---
title: "Pet Adoption Rate"
author: "Ashley Kim"
date: "4/28/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(caret)
library(tidyverse)
library(class)
library(plotly)
library(mice)
library(MLmetrics)
library(mltools)
library(data.table)

library(rio)
library(plyr)
library(tidyverse)
library(rpart)
library(psych)
library(pROC)
library(rpart.plot)
library(rattle)
library(mlbench)
```

## Questions 

How might we promote happiness with pet ownership? What factors impact the adoption of a pet, and likewise, what factors impact an animal not getting adopted?

## Background Information

Pets have been proven to have a positive effect on human happiness and mental health. By exploring adoption rates of pets we hope to find ways to promote happiness with pet ownership, as well as discovering what characteristics of an animal (age, color, intake condition, etc.) affect whether it'll be adopted or not using animal shelter data from Long Beach, California.

## Exploratory Data Analysis

```{r}

```

## Cleaning Dataset

```{r}
shelter_data <-  read_csv("/Users/ashleykim/Documents/Y3 2021-2022/DS3001Final/animal-shelter-intakes-and-outcomes.csv")
(column_index <- tibble(colnames(shelter_data)))

mice::md.pattern(shelter_data)
shelter_data <- shelter_data[,-c(1,2,5,7,12,13,15,18,19,20,21)] 
shelter_data<- na.omit(shelter_data)

shelter_data$days_in_shelter <- as.numeric(difftime(date_data$`Outcome Date`, date_data$`Intake Date`, units = "days"))

str(shelter_data)

mice::md.pattern(shelter_data)
```


### RESCUE VS NOT ADOPTED
```{r}
rescue_data <-shelter_data[!(shelter_data$`Outcome Type` == "ADOPTION" | shelter_data$`Outcome Type` =="FOSTER TO ADOPT"),]

rescue_data$`Outcome Type` <- fct_collapse(rescue_data$`Outcome Type`, 
                           Rescued = c("RESCUE"),
                           Not_Adopted = c("COMMUNITY CAT", "DIED", "DISPOSAL", "DUPLICATE", "EUTHANASIA", "FOSTER", "MISSING", "RETURN TO OWNER", 
                                  "RETURN TO RESCUE", "RETURN TO WILD HABITAT", "SHELTER, NEUTER, RETURN", "TRANSFER", "TRANSPORT", 
                                  "TRAP, NEUTER, RELEASE"))

rescue_data$`Animal Type` <- fct_collapse(rescue_data$`Animal Type`, 
                           Other = c("BIRD", "GUINEA PIG", "LIVESTOCK", "OTHER", "RABBIT", "REPTILE", "WILD"),
                           Cat = c("CAT"),
                           Dog = c("DOG"))

rescue_data$`Primary Color` <- fct_collapse(rescue_data$`Primary Color`, 
                           Other = c("APRICOT", "BC LYNX PT", "BL BRINDLE", "BL LYNX PT", "BLONDE", "BLUE","BLUE BRIND", "BLUE CREAM", "BLUE MERLE",
                                     "BLUE PT", "BLUE TABBY", "BLUE TICK", "BUFF", "C-T PT", "CALICO", "CALICO DIL", "CALICO PT", "CALICO TAB", 
                                     "CR LYNX PT", "CREAM", "CREAM PT", "CRM TABBY", "CRM TIGER", "DAPPLE", "FAWN", "FLAME PT", "GOLD", "GREEN",
                                     "L-C PT", "LC LYNX PT", "LI LYNX PT", "LILAC PT", "LIVER", "LIVER TICK", "LYNX PT", "ORANGE", "ORG TABBY", 
                                     "ORG TIGER", "PEACH", "PINK", "RD LYNX PT", "RED", "RED MERLE", "RUDDY", "S-T PT", "SABLE", "SEAL", "SEAL PT",
                                     "SILVER", "SL LYNX PT", "SLVR TABBY", "SNOWSHOE", "ST LYNX PT", "TORBI", "TORTIE", "TORTIE DIL", "TORTIE MUT",
                                     "TORTIE PT", "TRICOLOR", "UNKNOWN", "WHEAT", "Y BRINDLE", "YELLOW"),
                           Black = c("BLACK","BLK SMOKE", "BLK TABBY", "BLK TIGER"),
                           Brown_Tan = c("BR BRINDLE", "BRN MERLE", "BRN TABBY", "BRN TIGER", "BROWN", "CHOC PT", "CHOCOLATE", "SEAL", "SEAL PT",
                                         "TAN"),
                           Gray_White = c("GRAY", "GRAY TABBY", "GRAY TIGER", "WHITE"))
 
rescue_data$`Intake Condition` <- fct_collapse(rescue_data$`Intake Condition`, 
                           Other = c("AGED", "UNDER AGE/WEIGHT", "LIVESTOCK", "OTHER", "RABBIT", "REPTILE", "WILD", "WELFARE SEIZURES"),
                           Behavior = c("BEHAVIOR  MILD", "BEHAVIOR  MODERATE", "BEHAVIOR  SEVERE", "FERAL", "FRACTIOUS"),
                           Ill_Injured = c("ILL MILD", "ILL MODERATETE", "ILL SEVERE", "INJURED  MILD", "INJURED  MODERATE", "INJURED  SEVERE"),
                           Normal = c("NORMAL"))

rescue_data$`Intake Type` <- fct_collapse(rescue_data$`Intake Type`, 
                           Owner_Surrender = c("OWNER SURRENDER"),
                           Stray = c("STRAY"),
                           Wildlife = c("WILDLIFE"),
                           Other = c("CONFISCATE", "Euthenasia Required", "FOSTER", "RETURN", "SAFE KEEP", "WELFARE SEIZED", "Adopted Animal Return",
                                     "QUARANTINE", "TRAP, NEUTER, RETURN"))
 
rescue_data$`Jurisdiction` <- fct_collapse(rescue_data$`Jurisdiction`, 
                           LA = c("LA CITY", "LA COUNTY", "SIGNAL HILL", "TORRANCE AC"),
                           OC = c("ORANGE CNTY", "SEAL BEACH", "CERRITOS", "IRVINE", "GARDEN GROVE", "LA HABRA", "LOS ALAMITOS", "SEAACA", 
                                  "WESTMINSTER"),
                           LB= c("LONG BEACH", "DISTRICT1", "DISTRICT2", "DISTRICT3", "DISTRICT4", "DISTRICT5", "DISTRICT6", "DISTRICT7", "DISTRICT8",
                                 "DISTRICT9"),
                           OOA = c("OUT OF AREA"))

```

```{r}
table(shelter_data$`Animal Type`) # bird, cat, dog, other
table(shelter_data$`Primary Color`) # black, brown/tan, gray/white, other
table(shelter_data$`Sex`) 
table(shelter_data$`Intake Condition`) # normal, injured, ill, behavior, other
table(shelter_data$`Intake Type`) # owner surrender, stray, wildlife, other
table(shelter_data$`Jurisdiction`) # LA, Long Beach, Orange County, Other
table(shelter_data$`Outcome Type`) # adopted/rescued, not adopted
```

```{r}
shelter_data$`Animal Type` <- fct_collapse(shelter_data$`Animal Type`, 
                           Other = c("BIRD", "GUINEA PIG", "LIVESTOCK", "OTHER", "RABBIT", "REPTILE", "WILD"),
                           Cat = c("CAT"),
                           Dog = c("DOG"))

shelter_data$`Primary Color` <- fct_collapse(shelter_data$`Primary Color`, 
                           Other = c("APRICOT", "BC LYNX PT", "BL BRINDLE", "BL LYNX PT", "BLONDE", "BLUE","BLUE BRIND", "BLUE CREAM", "BLUE MERLE",
                                     "BLUE PT", "BLUE TABBY", "BLUE TICK", "BUFF", "C-T PT", "CALICO", "CALICO DIL", "CALICO PT", "CALICO TAB", 
                                     "CR LYNX PT", "CREAM", "CREAM PT", "CRM TABBY", "CRM TIGER", "DAPPLE", "FAWN", "FLAME PT", "GOLD", "GREEN",
                                     "L-C PT", "LC LYNX PT", "LI LYNX PT", "LILAC PT", "LIVER", "LIVER TICK", "LYNX PT", "ORANGE", "ORG TABBY", 
                                     "ORG TIGER", "PEACH", "PINK", "RD LYNX PT", "RED", "RED MERLE", "RUDDY", "S-T PT", "SABLE", "SEAL", "SEAL PT",
                                     "SILVER", "SL LYNX PT", "SLVR TABBY", "SNOWSHOE", "ST LYNX PT", "TORBI", "TORTIE", "TORTIE DIL", "TORTIE MUT",
                                     "TORTIE PT", "TRICOLOR", "UNKNOWN", "WHEAT", "Y BRINDLE", "YELLOW"),
                           Black = c("BLACK","BLK SMOKE", "BLK TABBY", "BLK TIGER"),
                           Brown_Tan = c("BR BRINDLE", "BRN MERLE", "BRN TABBY", "BRN TIGER", "BROWN", "CHOC PT", "CHOCOLATE", "SEAL", "SEAL PT",
                                         "TAN"),
                           Gray_White = c("GRAY", "GRAY TABBY", "GRAY TIGER", "WHITE"))
 
shelter_data$`Intake Condition` <- fct_collapse(shelter_data$`Intake Condition`, 
                           Other = c("AGED", "UNDER AGE/WEIGHT", "LIVESTOCK", "OTHER", "RABBIT", "REPTILE", "WILD", "WELFARE SEIZURES"),
                           Behavior = c("BEHAVIOR  MILD", "BEHAVIOR  MODERATE", "BEHAVIOR  SEVERE", "FERAL", "FRACTIOUS"),
                           Ill_Injured = c("ILL MILD", "ILL MODERATETE", "ILL SEVERE", "INJURED  MILD", "INJURED  MODERATE", "INJURED  SEVERE"),
                           Normal = c("NORMAL"))

shelter_data$`Intake Type` <- fct_collapse(shelter_data$`Intake Type`, 
                           Owner_Surrender = c("OWNER SURRENDER"),
                           Stray = c("STRAY"),
                           Wildlife = c("WILDLIFE"),
                           Other = c("CONFISCATE", "Euthenasia Required", "FOSTER", "RETURN", "SAFE KEEP", "WELFARE SEIZED", "Adopted Animal Return",
                                     "QUARANTINE", "TRAP, NEUTER, RETURN"))
 
shelter_data$`Jurisdiction` <- fct_collapse(shelter_data$`Jurisdiction`, 
                           LA = c("LA CITY", "LA COUNTY", "SIGNAL HILL", "TORRANCE AC"),
                           OC = c("ORANGE CNTY", "SEAL BEACH", "CERRITOS", "IRVINE", "GARDEN GROVE", "LA HABRA", "LOS ALAMITOS", "SEAACA", 
                                  "WESTMINSTER"),
                           LB= c("LONG BEACH", "DISTRICT1", "DISTRICT2", "DISTRICT3", "DISTRICT4", "DISTRICT5", "DISTRICT6", "DISTRICT7", "DISTRICT8",
                                 "DISTRICT9"),
                           OOA = c("OUT OF AREA"))

shelter_data$`Outcome Type` <- fct_collapse(shelter_data$`Outcome Type`, 
                           Adopted = c("ADOPTION", "FOSTER TO ADOPT"),
                           Not_Adopted = c("COMMUNITY CAT", "DIED", "DISPOSAL", "DUPLICATE", "EUTHANASIA", "FOSTER", "MISSING", "RETURN TO OWNER", 
                                  "RETURN TO RESCUE", "RETURN TO WILD HABITAT", "SHELTER, NEUTER, RETURN", "TRANSFER", "TRANSPORT", 
                                  "TRAP, NEUTER, RELEASE", "RESCUE"))

# 
```


# drop 
rescued vs not adopted
select if shelter outcome is adopted

```{r}
shelter_data[,c(3)] <- lapply(shelter_data[,c(3)], as.factor) # converting characters to factors
shelter_data2 <-shelter_data

normalize <- function(x){
 (x - min(x)) / (max(x) - min(x))
}

abc <- names(select_if(shelter_data, is.numeric))

shelter_data2[abc] <- lapply(shelter_data[abc], normalize)
```

```{r}
# one hot encoding
ab_c <- names(select_if(shelter_data2, is.factor))
shelter_data2 <- one_hot(as.data.table(shelter_data),cols=ab_c,sparsifyNAs = TRUE,naCols = FALSE,dropCols = TRUE,dropUnusedLevels = TRUE)

shelter_data2 <- shelter_data2[,-27]
```

## Methods

Using KNN and decision trees

Advantages of using KNN
- easy to interpret and naturally handles multiclass datasets 
- non-parametric

```{r}
# Running kNN algorithm 
table(shelter_data2$`Outcome Type_Adopted_Rescued`)[2]/sum(shelter_data2$`Outcome Type_Adopted_Rescued`)
# at random, we have an 100% of correctly picking if a dog is adopted (???????)

part_index_1 <- createDataPartition(shelter_data2$`Outcome Type_Adopted_Rescued`,
                                           times=1,
                                           p = 0.60,
                                           groups=1,
                                           list=FALSE)

train <- shelter_data2[part_index_1,]
tune_and_test <- shelter_data2[-part_index_1, ]

tune_and_test_index <- createDataPartition(tune_and_test$`Outcome Type_Adopted_Rescued`,
                                           p = .5,
                                           list = FALSE,
                                           times = 1)

tune <- tune_and_test[tune_and_test_index, ]
test <- tune_and_test[-tune_and_test_index, ]

dim(train)
dim(tune)
dim(test)
```

### KNN
```{r}
# training the classifier for k = 9 

set.seed(1984) # for randomized algorithm
shelter_9NN <- knn(train = train,#<- training set cases
               test = tune,    #<- tune set cases
               cl = train$`Outcome Type_Adopted_Rescued`,#<- category for true classification
               k = 9,#<- number of neighbors considered
               use.all = TRUE,
               prob = TRUE)# provides the output in probabilities 

str(shelter_9NN)
table(shelter_9NN)
table(tune$`Outcome Type_Adopted_Rescued`)

shelter_9NN # output

View(as.tibble(shelter_9NN))
View(as.tibble(attr(shelter_9NN,"prob")))
```

```{r}
# looking at how the kNN classification compares to the true class using the confusion matrix

kNN_res = table(shelter_9NN,
                tune$`Outcome Type_Adopted_Rescued`)
kNN_res
sum(kNN_res)  #<- the total is all the test examples

(2082+3154)/(2082+3154+0+0) # accuracy = TP+TN/(TP+TN+FP+FN)

kNN_res[row(kNN_res) == col(kNN_res)] # selecting true positives and true negatives 

kNN_acc = sum(kNN_res[row(kNN_res) == col(kNN_res)]) / sum(kNN_res) # accuracy rate calculation
kNN_acc 

# an 100% accuracy rate... find base rate to see the chance of guessing right if we don't know anything about the pet being adopted

confusionMatrix(as.factor(shelter_9NN), as.factor(tune$`Outcome Type_Adopted_Rescued`), positive = "1", dnn=c("Prediction", "Actual"), mode = "sens_spec")

#sensitivity, recall and true poss rate = TP/TP+FN
#specificity, true negative rate = TN/TN+FP
```

### Decision Tree - Adoption
```{r}
shelter_data <- shelter_data[,-c(5,8)]

prevalence <- 1-table(shelter_data$`Outcome Type`)[[1]]/length(shelter_data$`Outcome Type`)  # calculate the proportion of salary that is the positive class 
prevalence
```

The prevalence is the proportion of the positive class within the target variable, in this case, the pets that are adopted from the shelter.  In this data set, the prevalence is roughly 59%.  This means that roughly 41% of the data is the negative class, or pets that are not adopted.  This metric can be used a baseline because a model that always predicts the negative class will be correct 41% of the time.  

```{r}
set.seed(1)
part_index_1 <- caret::createDataPartition(shelter_data$`Outcome Type`,  # split the data with a .7 probability so that 70% of the data is chosen
                                           times=1,
                                           p = 0.70,
                                           groups=1,
                                           list=FALSE)

train <- shelter_data[part_index_1, ]  # subset the 70% chosen in the first partition into the train set
tune_and_test <- shelter_data[-part_index_1, ]  # subset the remaining 7 in a tune and test set 

set.seed(1)
tune_and_test_index <- createDataPartition(tune_and_test$`Outcome Type`,  # now split the tune and test set 50-50
                                           p = .5,
                                           list = FALSE,
                                           times = 1)

tune <- tune_and_test[tune_and_test_index, ]  # subset the 50% chosen into the tune set
test <- tune_and_test[-tune_and_test_index, ]  # subset the remaining 50% into the test set

dim(train)
dim(tune)
dim(test)
```

```{r}
features <- train[,c(-8)] # dropping salary column to get just the explanatory variables in features
target <- train$`Outcome Type`  # add just the salary column to target 


fitControl <- trainControl(method = "repeatedcv",  # use repeated cross validation with 5 folds and 3 repeats
                          number = 5,
                          repeats = 3, 
                          returnResamp="all",
                          classProbs = TRUE,
                          allowParallel = TRUE) 

tree.grid <- expand.grid(maxdepth=c(5,7,9,11))

set.seed(1984)  # set seed for reproducibility
shelter_mdl <- train(x=features,
                y=target,
                method="rpart2",#type of model uses maxdepth to select a model
                trControl=fitControl,#previously created
                tuneGrid=tree.grid,#expanded grid
                metric="ROC")#selected on of the metrics available from two variable summary.

shelter_mdl

plot(shelter_mdl)
varImp(shelter_mdl)
```

Variables with greatest importance are age, intake condition, animal type, intake type, and sex.

```{r}
rpart.plot(shelter_mdl$finalModel, type=4,extra=101)
shelter_mdl$results
```

### Decision Tree - Rescue
```{r}
rescue_data <- rescue_data[,-c(5,8)]

prevalence <- 1-table(rescue_data$`Outcome Type`)[[1]]/length(rescue_data$`Outcome Type`)  
prevalence
```

```{r}
set.seed(1)
part_index_1 <- caret::createDataPartition(rescue_data$`Outcome Type`,  # split the data with a .7 probability so that 70% of the data is chosen
                                           times=1,
                                           p = 0.70,
                                           groups=1,
                                           list=FALSE)

train <- rescue_data[part_index_1, ]  # subset the 70% chosen in the first partition into the train set
tune_and_test <- rescue_data[-part_index_1, ]  # subset the remaining 7 in a tune and test set 

set.seed(1)
tune_and_test_index <- createDataPartition(tune_and_test$`Outcome Type`,  # now split the tune and test set 50-50
                                           p = .5,
                                           list = FALSE,
                                           times = 1)

tune <- tune_and_test[tune_and_test_index, ]  # subset the 50% chosen into the tune set
test <- tune_and_test[-tune_and_test_index, ]  # subset the remaining 50% into the test set

dim(train)
dim(tune)
dim(test)
```

```{r}
features <- train[,c(-8)] # dropping outcome type column to get just the explanatory variables in features
target <- train$`Outcome Type`  


fitControl <- trainControl(method = "repeatedcv",  # use repeated cross validation with 5 folds and 3 repeats
                          number = 5,
                          repeats = 3, 
                          returnResamp="all",
                          classProbs = TRUE,
                          allowParallel = TRUE) 

tree.grid <- expand.grid(maxdepth=c(5,7,9,11))

set.seed(1984)  # set seed for reproducibility
rescue_mdl <- train(x=features,
                y=target,
                method="rpart2",#type of model uses maxdepth to select a model
                trControl=fitControl,#previously created
                tuneGrid=tree.grid,#expanded grid
                metric="ROC")#selected on of the metrics available from two variable summary.

rescue_mdl

plot(rescue_mdl)
varImp(rescue_mdl)
```

```{r}
rpart.plot(rescue_mdl$finalModel, type=4,extra=101)
rescue_mdl$results
```

## Evaluation of our model

- adopted animals tend to be younger, owner-surrendered (not strays) and of the cat/dog variety

- often wildlife, older animals and those who come in with behavioral issues and illness/injuries are not adopted as well 


## Conclusion

Shelters should focus on promoting animals in their shelters that are typically older 
Our main priority for this exploration is to see which factors contribute the most to an animal not being adopted in order to use this data to promote more instances of adoptions in shelters. This can be done through the shelter marketing older animals that are typically strays and with health issues more frequently. But a select quota of pets that hold a bunch of these "unadoptable" characteristics  may continue to not be adopted at all despite a higher amount of promotion. However, there are different ways to handle these animals without them going to a home. Other opportunities for these animals lie in rescues, animal sanctuaries, and more. By doing this, it can help human society in a different way by people coming through and visiting and also the opportunity to help these animals through volunteer work as well.

## Future work
- maybe more quantitative variables? were limited to only several methods, wanted to try clustering but didn't have enough quantitative variables (i.e. weight)

- possibility  of overfitting  with KNN(at least on my end)